```{r echo=F, message=F, warning=F}
rm(list=ls())
```

# Introduction to R

This chapter provides an introduction to R, covering basic
concepts like defining variables, performing mathematical
operations, and using built-in functions. It also introduces
various data types (numeric, character, logical, and factor)
and containers (vector, list, and data frame) in R, along
with how to create and access their elements. It also covers
control flow with for and while loops, conditional flow with
if statements, and the creation of custom functions. Lastly,
it discusses inspecting objects using functions like ls(),
rm(), class(), str(), head(), tail(), and summary() to
understand and manage the R environment.

## Basics
```{r}
# Use R as a calculator
2 + 2

# Store results in variables
x <- 2 + 2
x

# Perform mathematical operations on variables
x <- 2 + 2
y <- 3 + 7
x + y
```

## Functions
- Functions are objects that take *arguments* as inputs,
  perform some operation on those inputs, and return the
  results.

- `ls()` is a function that reports what objects (e.g.,
  variables you have defined) are in the current environment
  and therefore available for you to interact with.

```{r}
ls()
```

- `rm()` is a function that can remove objects from the
  current environment.

- It takes several arguments. Type `?rm()` to see help
  information on how to use it.

- We will commonly combine `ls()` and `rm()` to remove all
  objects from the current environment. This is a good thing
  to do at the beginning of every new script you write.

```{r}
# Remove all objects from the current R session.
rm(list = ls())
```

## Data types

### `numeric`
```{r}
x <- 2
class(x)
```

### `character`
```{r}
x <- 'a'
class(x)
```

### `logical`
```{r}
x <- TRUE
class(x)
```

### `factor`

- A `factor` is a *categorical* data type. They are most
  often used to code for different conditions in an
  experiment.

```{r}
x <- factor(1)
class(x)
```

## Containers

### `vector`

- Create a vector with the function `c()`.

- The elements of a `vector` must be of the same type.

- Access element `i` of `vector` `x` with square brackets
  (e.g., `x[i]`)

```{r}
# Create a vector with any three numbers you like.
x <- c(1, 2, 3.14159)
x

# Access the third element of x
x3 <- x[3]
x3
```

### `list`

- Create a list with the function `list()`

- The elements of a `list` can be of different types.

- Access element `i` of `list` `x` with *double* square
  brackets (e.g., `x[[i]]`)

```{r}
# Create a three element list containing one numeric
# item, one `character` item, and one logical item.
x <- list(3.14159, 'pi', TRUE)
x

# Access the third element of x
x3 <- x[[3]]
x3
```

### `data.frame`

- Create a `data.frame` with the function `data.frame()`

-`data.frame` is pretty close what you might think of as an
excel spreadsheet.

- Access column `x` in `data.frame` `df` with the `$`
  operator (e.g., `df$x`).

```{r}
# Create vectors to later store in a data frame
x <- c('I', 'I', 'I', 'II', 'II', 'II', 'III', 'III', 'III', 'IV', 'IV', 'IV')
y <- c('a', 'a', 'b', 'b', 'c', 'c', 'd', 'd', 'e', 'e', 'f', 'f')
z <- rnorm(12)

# Create the data frame
df <- data.frame(x, y, z)
df

# Access column x
df$x
```

## Loops

### `for` loops

- `for` loops will run a chunk of code repeatedly for a
  fixed number of iterations.

- The general syntax of a `for` loop is as follows:

```{r}
for(x in y) {
  # On the first iteration, `x` will take the value `y[1]`
  # On the second iteration, `x` will take the value `y[2]`
  # On the third iteration, `x` will take the value `y[3]`
  # The loop will end after `x` has taken the value `y[length(y)]`
  # That is, the loop will end when we have iterated through
  # all elements in `y`
}
```

- As an example suppose we want to print the numbers 1, 2, 3

```{r}
# dumb monkey way to print 1, 2, 3
print(1)
print(2)
print(3)

# smart monkey way to print 1, 2, 3
for(i in c(1, 2, 3)) {
  print(i)
}
```

### `while` loops

- `while` loops will run a chunk of code repeatedly over and
  over again until some `logical` condition is met.

- You have to be careful with these, because if your code
  never sets up a stopping condition, then the loop will
  execute until your computer turns to dust.

- The general syntax of a `for` loop is as follows:

```{r, eval=F}
condition <- TRUE
while(condition) {
  # On the first iteration, `x` will take the value `y[1]`
  # On the second iteration, `x` will take the value `y[2]`
  # On the third iteration, `x` will take the value `y[3]`
  # The loop will end only when `condition` is set to `FALSE`
}
```

- Lets again consider the example printing the numbers 1, 2,
  3

```{r}
# dumb monkey way to print 1, 2, 3
print(1)
print(2)
print(3)

# smart monkey way to print 1, 2, 3
x <- 1
while(x < 4) {
  print(x)
  x <- x + 1 # without this line the loop would run forever
}
```

## Conditional flow

- Very often we will want to execute a chunk of code only in
  some situations (e.g., for a particular experiment
  condition) and we will want to run some other chunk of
  code in other situations.

- The primary method for doing this is to use `if`
  statements. The general syntax of an `if` statement is as
  follows:

```{r, eval=F}
if(condition) {
  # code to run if condition is TRUE
} else {
  # code to run if condition is FALSE
}
```

* For example, suppose we want to print whether or not a
  number is less than 5.

```{r}
x <- 3
if(x < 5) {
  print('x is less than 5')
} else {
  print('x is greater than or equal to 5')
}
```

## Custom functions

- Custom functions are very useful because they allow us to
  flexibly reuse the same chunk of code in different places
  without having to rewrite the entire chunk.

- The general syntax for defining functions is as follows:

```{r}
function_name <- function(argument_1, argument_2, ...) {
  ## code to run when the function is called. Can use
  ## `argument_1`, `argument_2`, and any other argument
  ## passed in.
  ##...
  return(the_result)
}
```

- `function_name` is the name of the function.

- `argument_1`, `argument_2`, etc. are variables that you
  want the code chunk inside the function to use.

- `the_result` is a variable that you will have to be
  careful to define in the code chunk in the function.

- Consider the following example:

```{r}
my_func <- function(x, y) {
  z <- x + y - 1
  return(z)
}

my_func(x=2, y=3)
```

- `my_func` take two arguments `x` and `y` and returns `x +
  y - 1`

## Inspect existing objects

- `R` has many built-in functions that are useful for
  inspecting what sort of thing an existing object is. We
  illustrate the use of some of these functions below.

```{r}
# define some variables (objects) to inspect
var_1 <- 10
var_2 <- "apple"
var_3 <- TRUE
var_4 <- c(1, 2, 3)
var_5 <- list('a', 'b', 'c')
var_6 <- data.frame(v4=var_4, v5=var_5)

# substitute different variables into the following
# functions to see how they help you inspect existing
# objects.
str(var_1)
class(var_1)
head(var_1)
tail(var_1)
summary(var_1)
```

## Summary

Here's a list of the R functions we saw in the above
examples:

1. `ls()` - Lists objects in the current environment.
2. `rm()` - Removes objects from the current environment.
3. `class()` - Returns the class (type) of an object.
4. `factor()` - Creates a factor (categorical data type).
5. `c()` - Combines values into a vector or list.
6. `list()` - Creates a list.
7. `data.frame()` - Creates a data frame.
8. `rnorm()` - Generates normally distributed random numbers.
9. `print()` - Prints its argument.
10. `str()` - Displays the structure of an R object.
11. `head()` - Returns the first parts of an object.
12. `tail()` - Returns the last parts of an object.
13. `summary()` - Provides a summary of an object's properties.

Additionally, we saw how to create and use custom functions,
as well as how to use control flow constructs like `for`
loops, `while` loops, and `if` statements. 

