---
title: "Tutorial 1"
author: "Author: Matthew J. Crossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    # code_folding: show
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

## Work through these practice exercises
* It's a good idea to work through these on your own, but if
  you get very stuck, solutions can be found
  [here](tutorial_1_solutions.html)

* Start by making sure you have a clean session.

```{r, echo=T}
# clean session
rm(list=ls())
```

### 1.
* a. Define a variable named `ans_1a_x` and set its value to
  `10`, a variable named `ans_1a_y` and set its value to
  `5`, and a variable named `ans_1a_z` and set its value to
  the sum of `ans_1a_x` and `ans_1a_y`. The line of code
  that you write to define `ans_1a_z` must include the `+`
  operator.

* b. Define a variable named `ans_1b_x` and set its value to
  `"10"`, a variable named `ans_1b_y` and set its value to
  `"5"`, and a variable named `ans_1b_z` and set its value
  to the sum of the *numeric values* of `ans_1b_x` and
  `ans_1b_y`. The line of code that you write to define
  `ans_1b_z` must include the `+` operator and the
  `as.numeric()` function.

* c. Define a variable named `ans_1c` and set its value to
  the `logical` result indicating if `ans_1a_x` is equal to
  `ans_1a_y`. The line of code that you write to define
  `ans_1c` must include the `==` operator.

### 2.
* a. Create a `vector` named `ans_2a` that contains the
  elements `2`, `3`, `5`, `7`, and `11`.

* b. Create a `list` named `ans_2b` that contains the
  elements `1`, `2`, `"a"`, `"b"` and `TRUE`.

* c. Create a `data.frame` named `ans_2c` that contains
  `v1`, `v2`, and `v3` as defined below as columns. Make
  sure the column names of the `data.frame` you create are
  equal to the variable names `v1`, `v2`, and `v3`.

```{r, echo=T}
v1 <- c('I', 'I', 'I', 'I', 'II', 'II', 'II', 'II')
v2 <- c('a', 'a', 'b', 'b', 'c', 'c', 'd', 'd')
v3 <- c(-1.6297880,
        -1.0738506,
         0.0299236,
        -1.5435811,
        -0.5133278,
        -1.4716107,
        -1.1986316,
        -1.5548207)
```

### 3.
* a. Write a line of code that returns the third element
  from the `vector` named `tmp` defined in the following
  code chunk and store the result in a variable named
  `ans_3a`.

```{r, echo=T}
tmp <- c(0.24, 0.015, 1.34, -1.00, -0.15)
```

* b. Write a line of code that returns the second element
  from the `list` named `tmp` defined in the following code
  chunk and store the result in a variable named `ans_3b`.
  Be sure that the data type of `ans_3b` is `character`.

```{r, echo=T}
tmp <- list(0.5, "W", FALSE)
```

* c. Write a line of code that returns the column named `v2`
  from the `data.frame` named `tmp` defined in the following
  code chunk and store the result in a variable named
  `ans_3c`. Be sure that the line of code that you write
  includes the `$` operator.

```{r, echo=T}
tmp <- data.frame(v1=c(1, 2, 3),
                  v2=c('A', 'B', 'C'),
                  v3=c(TRUE, TRUE, FALSE))
```

### 4.
* a. For the following code chunk, what is the data type of
  the second element of `tmp`? Store your answer in a
  variable named `ans_4a` by copying, pasting, and
  uncommenting one of the commented out lines of code below.

```{r, echo=T}
element_1 <- "A"
element_2 <- 2
element_3 <- FALSE
tmp <- list(element_1, element_2, element_3)

# ans_4a <- "character"
# ans_4a <- "numeric"
# ans_4a <- "logical"
```

* b. For the following code chunk, what is the data type of
  the second element of `tmp`? Store your answer in a
  variable named `ans_4b` by copying, pasting, and
  uncommenting one of the commented out lines of code below.

```{r, echo=T}
element_1 <- "A"
element_2 <- 2
element_3 <- FALSE
tmp <- c(element_1, element_2, element_3)

# ans_4b <- "character"
# ans_4b <- "numeric"
# ans_4b <- "logical"
```

### 5.
* The function `generate_experiment_result()` defined in the
  following code chunk takes `sample_size` as an argument
  and returns the results from a silly experiment in which
  we sample from a normal distribution.

```{r, echo=T}
set.seed(0)
generate_experiment_result <- function(sample_size) {
  result <- rnorm(sample_size)
  return(result)
}

experiment_result <- generate_experiment_result(10)

experiment_result_mean <- mean(experiment_result)

critical_value <- 1.5
if(experiment_result_mean > critical_value) {
   experiment_decision <- TRUE
} else {
   experiment_decision <- FALSE
}
```

* a. Is the function `generate_experiment_result ` a
  **descriptive statistic**, an **inferential statistic**, a
  **random variable**, or a **sample** from a random
  variable?  Please use one of the following lines of code
  to save your answer.

* b. Is the variable `experiment_result` a **descriptive
  statistic**, an **inferential statistic**, a **random
  variable**, or a **sample** from a random variable?
  Please use one of the following lines of code to save your
  answer.

* c. Is the variable `experiment_result_mean` a
  **descriptive statistic**, an **inferential statistic**, a
  **random variable**, or a **sample** from a random
  variable?  Please use one of the following lines of code
  to save your answer.

* d. Is the variable `experiment_decision` a **descriptive
  statistic**, an **inferential statistic**, a **random
  variable**, or a **sample** from a random variable?
  Please use one of the following lines of code to save your
  answer.

### 6.
* a. Consider the following code chunk:

```{r, echo=T, eval=F}
x <- 5
for(i in c(1, 3, 5)) {
  x <- i + 2
}
```

* What is the value of `x` after the loop has finished
  executing?
* What is the value of `i` after the loop has finished
  executing?
* Modify the code chunk such that the final value of `x` is
  10 and `i` is 20.

* b. Consider the following code chunk:

```{r, echo=T, eval=F}
x <- 10
y <- c(1, 2, 3)
while(x > 5) {
  y <- c(x, y)
  x <- x - 1
}
```

* How many elements does `y` contain after the loop has
  finished?
* What is the value of the final element in `y` and can you
  figure this out without executing the code?

* c. Consider the following code chunk:

```{r, echo=T, eval=F}
if(x > y) {
  z <- 10
} else {
  z <- 2
}
```

* What is the value of `z` if `x <- 10` and `y <- -10`
* What is the value of `z` if `x <- -5` and `y <- 5`
* What is the value of `z` if `x <- 0` and `y <- 0`

* d. Consider the following code chunk:

```{r, echo=T, eval=F}
for(i in 1:10) {
  x <- i * 2
  if(x > 5) {
    break
  }
}
```

* How many times will this loop run?
* What is the value of `i` and `x` when the loop stops?

* e. consider the following code chunk:

```{r, echo=T, eval=F}
if(x == y) {
  z <- 1
} else if(x > y) {
  z <- 2
} else if(x < y) {
  z <- 3
}
```

* What is the value of `z` if `x <- 10` and `y <- -10`
* What is the value of `z` if `x <- -5` and `y <- 5`
* What is the value of `z` if `x <- 0` and `y <- 0`

### 7.

* a. Consider the following code chunk:

```{r, echo=T, eval=F}
f <- function(x) {
  y <- x^2
  return(y)
}
z <- f(3)
```

* What are the values of `x`, `y` and `z`?

* b. Consider the following code chunk:

```{r, echo=T, evsl=F}
f <- function(x) {
  y <- x^2
  return(y)
}

g <- function(x) {
  y <- x - 2
  return(y)
}

z <- f(g(4))
```

* What are the values of `x`, `y` and `z`?

* c. Consider the following code chunk:

```{r, echo=T, eval=F}
f <- function(x, y, z) {
  res <- x + y - z
  return(res)
}
```

* Is `f(2, 1, -4)` a valid way to call the function `f`?
* Is `f(y=2, 1, -4)` a valid way to call the function `f`?
* Is `f(y=2, 1, -4)` a *wise* way to call the function `f`?
* What is the wisest way to call the function `f`?
