---
title: "Tutorial 3 - Visualise data with `ggplot2`"
author: "Author: Matthew J. Crossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    # code_folding: show
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, collapse=T)
```

```{r, echo=F}
library(data.table)
library(ggplot2)
rm(list=ls())
```

## Work through these practice exercises
* It's a good idea to work through these on your own, but if
you get very stuck, solutions can be found
[here](tutorial_3_solutions.html)

* These practice problems use the `diamonds` data set that is loaded
automatically when you load `ggplot2`. By default it exists
as a `data.frame`. You can convert it to a `data.table` with
the following line:

```{r, echo=F}
d <- as.data.table(diamonds)
```

* For each practice problem, replicate each `ggplot` figure
and `data.table` output shown below **exactly**.

## 1.

```{r, echo=F}
ggplot(d[cut %in% c('Ideal', 'Premium')], aes(carat, fill=as.factor(cut))) +
  geom_histogram(alpha=0.75)

d[cut == 'Ideal', .(mean_carat=mean(carat),
        range_carat=diff(range(carat))),
    .(cut)]
```

## 2.

```{r, echo=F}
ggplot(d[clarity %in% c('SI2', 'SI1', 'VS1')], 
       aes(cut, fill=clarity)) +
  geom_bar(position=position_dodge())

d[, .N, .(cut, clarity)]
```

## 3.

```{r, echo=F}
ggplot(d[price < 10000], aes(cut, price, fill=clarity)) +
  geom_boxplot()

d[price < 1000, .(median_cty=as.numeric(median(price)),
        range_cty=diff(range(price))),
    .(cut, clarity)]
```

## 4.

```{r, echo=F}
ggplot(d[color == 'E'], aes(carat, price, colour=cut, fill=color)) +
  geom_point()
```
