## Computing probabilities from probability distributions
With discrete probability distributions, we were able to
compute the probability of certain events by counting the
total number of ways for an outcome to occur and dividing by
the total number of possible outcomes.  However, for
*continuous* probability distributions, this method doesn't
work. To see this, consider the continuous distribution
associated with the simple reaction time experiment in the
previous section.

Suppose we are interested in $p(X=1)$.  We first need to
count the number of ways for an outcome to occur. For
$p(X=1)$ -- a single outcome --  there is just one way. Next
-- assuming equal probability of all outcomes (not a good
assumption in this case but this is the least of our
troubles at the moment) -- we need to divde by the total
number of possible outcomes.  However, since $X$ is
continuous, there are an infinite number of possible
outcomes. So we are left dividing one by infinity. Any real
number divided by infinity is zero. So here lies our
problem. The probability of any single outcome in a
continuous distribution is zero.

Perhaps the situation improves if instead of asking about
the probbaility of single outcomes, we ask about the
probability of ranges of outcomes. For example, what is the
probability that $X$ is less than $1$? How many ways are
there for $x$ to be less than $1$? Since $x$ is continuous,
there are *infinite* ways for this occur! What about the
total number of possible outcomes? Right. Infinity again. So
we are left dividing one infinity by another infinity. Quite
a confusing and ill-defined situation.

### Discrete distributions
With discrete probability distributions -- also called
probability mass functions (pmfs) -- finite and well-defined
probabilities exist for every single possible outcome.
**This means that you can plot the distribution, and read
probabilities right off the y-axis of the plot.**

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)

rm(list=ls())

x <- seq(0, 15, 1)
fx <- dbinom(x, 15, 0.5)
d <- data.table(x=x, fx=fx)
g1 <- ggplot(d, aes(x, fx)) +
    geom_point() +
    ylab('Probability') +
    theme(
        axis.text=element_text(size=18),
        axis.title=element_text(size=24)
    )

x <- seq(0, 15, 1)
fx <- dbinom(x, 15, 0.5)
d <- data.table(x=x, fx=fx)
d[, px := x < 8]
g2 <- ggplot(d, aes(x,
              fx,
              colour=factor(px),
              fill=factor(px))) +
    geom_point() +
    geom_segment(data=d[px==TRUE],
                 aes(x=x,
                     xend=x,
                     yend=fx,
                     y=0)) +
    ylab('Probability') +
    # $$p(X<8) = \sum_{i=0}^7 p(x_{i})$$
    ggtitle('Probability of X < 8') +
    theme(
        axis.text=element_text(size=18),
        axis.title=element_text(size=24),
        legend.position='none'
    )
ggarrange(g1, g2, ncol=2)
```

### Continuous distributions
With continuous probability distributions -- aslo called
probability density functions (pdfs) -- the probability of
any single outcome is zero. The distribution represents
*probability density* rather than probability. **In this
case, probability is computed by taking the area under the
curve.**

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)

rm(list=ls())

x <- seq(-5, 5, 0.01)
fx <- dnorm(x, 0, 1)
d <- data.table(x=x, fx=fx)

g1 <- ggplot(d, aes(x, fx)) +
    geom_line() +
    ylab('Probability Density') +
    theme(
        axis.text=element_text(size=18),
        axis.title=element_text(size=24)
    )

x <- seq(-5, 5, 0.01)
fx <- dnorm(x, 0, 1)
d <- data.table(x=x, fx=fx)
d[, px := x < 1]

g2 <- ggplot(d, aes(x,
              fx,
              colour=factor(px),
              fill=factor(px))) +
    geom_line() +
    geom_ribbon(data=d[px==TRUE],
                aes(x=x,
                    ymin=0,
                    ymax=fx),
                alpha=0.3) +
    ylab('Probability Density') +
    # $$ p(X<1) = \int_{-\inf}^{1} f(x) dx $$
    ggtitle('Probability of X < 1') +
    theme(
        axis.text=element_text(size=18),
        axis.title=element_text(size=24),
        legend.position='none'
    )
ggarrange(g1, g2, ncol=2)
```
