```{r echo=F, message=F, warning=F}
rm(list=ls())
```
# Normal Distribution

## Learning objectives

* Define and understand **Normal** random variables.

## Normal distribution

### Definition
The normal distribution is defined by two parameters:

- $\mu$: population mean
- $\sigma$: population standard deviation

It is just a quirk of the normal distribution that the
parameters that define it correspond to its mean and
standard deviation. This will not generally be true for most
other distributions that we will deal with.

### Probability distribution

The normal distribution is continuous, and the continuous
probability distributions are called probability density
functions (pdfs).

The following equation defines the pdf of a normal
distribution:

$$
f(x) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{1}{2}{\frac{(x - \mu)^2}{\sigma}}}
$$

#### Normal distribution: function of $\mu$ and $\sigma$

Lets have a look at how the normal distribution changes with
different mean and variance.

```{r, echo=F}
## Lets more formally examine the of effect different values of n and p
mean_levels <- c(-25, 0, 25)
var_levels <- c(1, 5, 10)

x <- seq(-50, 50, 0.1)

## Watch out: `for()` loops ahead!
mean_rec <-  c()
var_rec <- c()
x_rec <- c()
p_norm_rec <- c()
for(mu in mean_levels) {
    for(sigma in var_levels) {
        mean_rec <- c(mean_rec, rep(mu, length(x)))
        var_rec <- c(var_rec, rep(sigma, length(x)))
        x_rec <- c(x_rec, x)
        p_norm_rec <- c(p_norm_rec, dnorm(x, mu, sigma))
    }
}

d <- data.table(mean=mean_rec, var=var_rec, x=x_rec, p_norm=p_norm_rec)

ggplot(d, aes(x=x, y=p_norm)) +
    geom_line() +
    xlab('X') +
    ylab('Probability Density') +
    facet_grid(mean~var)
```

#### The standard normal (Z)

The *standard normal* distribution is a normal distribution
with zero mean and unit variance ($\mu=0$, $\sigma^2=1$). It
is also called the $Z$ distribution. It looks like this:

```{r, echo=F}

mu <- 0
sigma <- 1

x <- seq(-5, 5, 0.01)
fx <- dnorm(x, mu, sigma)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_line()

```
