## Probability distributions
If $X$ is a random variable, then the probability
distribution of $X$ is a function that gives the probability
that $X$ will yield on a particular value when it is
sampled. It is very imoprtant to emphasise that a
probability distribution specifies the probability of every
possible outcome the random variable can take.

### Example: Discrete random variable
Lets return to our coin flip experiment in which we flip a
coin twice. The sample space for the experiment is given by
the following table:

| Outcome # | 1st Flip | 2nd Flip |
|-----------|----------|----------|
| 1         | H        | H        |
| 2         | H        | T        |   
| 3         | T        | H        |
| 4         | T        | T        |

Now, let $X$ be the random variable that counts the number
of heads in the experiment. The probability distribution of
$X$ is given by the following table:

| $x$ | $P(X=x)$ |
|-----|----------|
| 0   | 0.25     |
| 1   | 0.5      |
| 2   | 0.25     |

Note that here we have assumed that coin is fair and that
the P(H) = P(T) = 0.5. With this assumption we are able to
compute the probability of each outcome in the sample space
by adding the number of heads in each row of the sample
space table and divding by the total number of rows. Also,
note that $X$ cannot yield fewer than zero heads or more
than two heads, so the probability distribution of $X$ is
fully specified by the above table. Lets plot the
probability distribution of $X$ and also draw a few samples
from $X$.

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)

rm(list=ls())

x <- c(0, 1, 2)
y <- c(0.25, 0.5, 0.25)
d <- data.table(x, y)
g1 <- ggplot(d, aes(x, y)) +
  geom_bar(stat='identity') +
  xlab('X=x') +
  ylab('P(X=x)') +
  theme(aspect.ratio = 1)

n <- 1000
x <- rbinom(n, 2, 0.5)
d <- data.table(x)
g2  <- ggplot(d, aes(x)) +
  geom_bar(stat='count') +
  xlab('X=x') +
  ylab('Count') +
  theme(aspect.ratio = 1)
ggarrange(g1, g2, ncol=2)
```

As in previous examples we see that the majority of the
samples are clustered around the outcomes associated with
the highest probability in the probability distribution.

### Example: Continuous random variable
Lets return to our reaction time experiment in which we have
a random variable $X$ that measures the time it takes for a
person to react to a stimulus. The sample space of this
experiment is the set of all real numbers greater than zero
which is both continuous and infinite. This means that we
cannot specify the probability of every possible outcome
using a table as we did for the discrete random variable.
We can howeber visualise it using a plot. Lets assume that
the probability distribution of $X$ is given by the
following plot:

```{r, echo=F}
library(data.table)
library(ggplot2)

rm(list=ls())

# define sample space as a continuous random variable
x <- seq(0, 10, 0.1)

# define probability distribution as F distribution
y <- df(x, 5, 2)

# create a data table
d <- data.table(x, y)

# plot the probability distribution
g1 <- ggplot(d, aes(x, y)) +
  geom_line() +
  xlab('X=x') +
  ylab('P(X=x)') +
  theme(aspect.ratio = 1)

# draw a few samples from X
n <- 1000
xs <- rf(1000, 5, 2)

# plot the samples
d <- data.table(xs)

g2 <- ggplot(d, aes(xs)) +
  geom_histogram(binwidth=0.1) +
  xlab('X=x') +
  ylab('Probability density f(x)') +
  theme(aspect.ratio = 1)

ggarrange(g1, g2, ncol=2)
```

We again see that the majority of the samples are clustered
around the outcomes associated with the highest probability.
However, note that there is an important twist here. Notice
that the y-axis of the plot of the probability distribution
is labelled *probability density f(x)* and not simply
*P(X=x)*.  **This is because when $X$ is continuous
$P(X=x)=0$ for all $x$.** The $f(x)$ label refers to
probability density. This is a very cool and
counterintuitive idea that has important implications for
how we interact with probability distributions. We turnt to
this next.
