```{r echo=F, message=F, warning=F}
rm(list=ls())
```

# Bernoulli random variables

## Learning objectives

* Define and understand **Bernoulli** random variables.

## Bernoulli random variable

### Definition
A single sample from a any random variable that produces a
dichotomous outcome is formally defined as a *Bernoulli*
trial. To be a Bernoulli trial, a few conditions must be
met:

(1) Each trial yields one of the two outcomes usually called
success ($S$) and failure ($F$).

(2) For each trial, the probability of success $P(S)$ is the
same and is denoted by $p = P(S)$. The probability of
failure is then $q = P(F) = 1 - P(S)$ for each trial.

(3) Trials are **independent**. The probability of success
in a trial does not change given any amount of information
about the outcomes of other trials.

Bernoulli random variables are completely defined by the
single parameter $p = P(S)$. If $X$ is a Bernoulli random
variable then we would write:

$X \sim Bernoulli(p)$

### Probability distribution
The probability distribution of a Bernoulli random variable
is very simple. We simply need to indicate two probabilities
as follows.

```{r, echo=F}
library(data.table)
library(ggplot2)

rm(list=ls())

# Specify the probability of success
p <- 0.3

# On the y-axis, we want to plot the probability of the
# possible outcomes (S and F)
y <- c(0.3, 0.7)

# On the x-axis, we want to indicate the outcome names
x <- c('S', 'F')

# To plot things with ggplot we need our data in a
# data.table
d <- data.table(x, y)

ggplot(d, aes(x, y)) +
  geom_segment(aes(xend=x, yend=0)) +
  xlab('X = Bernoulli') +
  ylab('Probability') +
  theme(aspect.ratio = 1)
```
